---
title: "BSCS_306_Practical"
author: "Dr Tahseen Jilani"
date: "1/14/2022"
output: word_document
---

# Selecting a dataset
# -------------------
Choose a data set from R built-in datasets or from UCI or Kaggle.

```{r}
data() # To see the available data sets.
```

You can write about the data set you have selected. About the variables, number of rows (samples), columns (variables).

Outside the cells, if you put # it will be converted into heading. ## or subheading and so on. Useful website: http://www.sthda.com/english/wiki/r-built-in-data-sets

```{r}
data(mtcars)  # 1. Loading 
head(mtcars) # 2. Print only top few values of the data
dim(mtcars) # Check the dimensions of the data.
#nrow(mtcars) # ncol(mtcars)

print(summary(mtcars))
# To print more than one results use, print(your command). 
# For example(print(dim(mtcars)))


#install.packages("readxl")
#library("readxl")

getwd()
#setwd("give\the\folder\path\here)
# mydata = read.csv("filename.csv")

library(ggplot2)

```

```{r}
# Example of taking summary of a dataset.
# iris = iris  # Loading iris flowers data from R built-in datasets.
mtcars=  mtcars


# Note: In case you need to convert a categorical variable into numerical categories. For example gender = {Male, Female} will become Gender_c = {1,0} (based on alphabetical order of F and M).

table(mtcars$cyl, mtcars$gear)

as.numeric(iris$Species)
# Now for the species column, we can use 
iris$Species = as.numeric(iris$Species) # Overwrite th actual names with numbers. 
# R is case sensitive. Take care of small and capital letters
```

# 3. Basic plots
# ---------------
It is better to make some basic plots and discuss any possible outlier or any unique behavior(s) of the data. You are free to make as many plots and comments.  
You can add scatter-plots, histograms, box-plots etc in your report as appropriate. 

```{r}
attach(mtcars)
# In place of mtcars$mpg jus use mpg.

boxplot(iris$Sepal.Length , iris$Species) # Simple box plot for the variable iri-Sepal length versus Species.

hist(mtcars$mpg, breaks= 15, col = 'Grey',
     xlab= 'Miles per gallon', 
     main = 'Histogram for miles per gallon for the Cars dataset')

boxplot(mtcars$hp ~ mtcars$gear, main="Fig.-1: Boxplot of Mileage of Four Types of gear", col= rainbow(3))

plot(mtcars$mpg, mtcars$hp, col = "Blue")

# you can use ggplot2 as well. But optional.
#install.packages(ggplot2)
library(ggplot2)

```

# Correlation
# -----------
```{r}
cor(mtcars)
# Look at the magnitude and sign of the values. Comment and see which variables are more closely related to each other.


# You can plot the correlation values using corr.plot in the corrplot Package.
# -----------------------------------------------------------------------------
library(corrplot)
cor.mat.mtcars = cor(mtcars)
corrplot(cor.mat.mtcars)

```



# Confidence interval
# -------------------
```{r}
# You can get help for this from the following links

# The commands can be found here.
# https://cran.r-project.org/web/packages/Rmisc/Rmisc.pdf
# You can make confidence interval for mean(s) and Variance(s).
  
#install.packages(Rmisc)
library(Rmisc)

#CI(x, ci = 0.95) # Calculates the confidence interval of a vector of data.
CI(mtcars$mpg, ci = 0.95) # Calculates the confidence interval of a vector of data.
# Also comment about whether the mean is significantly different from zero.

# This can be done manually
# https://www.r-bloggers.com/2021/04/calculating-confidence-interval-in-r/
# https://www.cyclismo.org/tutorial/R/confidence.html

```


# Hypothesis testing
# -------------------
```{r}
# In the stat pacakge, you can find different hypothsis tests.
library(stats)

# One sample t-test
# -------------------
t.test(mtcars$mpg, mu = 20) # to test: Is the mean value of mpg differ from 20 or not?
t.test(mtcars$mpg, mu =20 , alternative = 'greater') # if you want to test on sided alternative. 

# Two sample t-test for two independent samples
# ----------------------------------------------
# t.test(iris$Sepal.Length, iris$Petal.Width)

# Two sample t-test for a variable with two categories defined in other variable.
# For example: Grades comparsion between male and female students. So we will have two variables 1. grades and 2. gender (with exactly two categories).
# -----------------------------------------------------------------------------------
t.test(mtcars$mpg ~ mtcars$vs , data = mtcars)

# Useful help for this task can be found at https://data-flair.training/blogs/hypothesis-testing-in-r/

```

# Chi-square test
# ---------------
```{r}
# Chose any two categorical variables (or some variable with up to 10 unique values).
# Then apply the Chi-Square test to see test of association/ independence.

chisq.test(iris$Sepal.Length, iris$Species)

chisq.test(mtcars$mpg, mtcars$gear)

# You will get the Chi-square test value. Also comment on the P-value of the test.
# you can add more than one Chi-square test to get more data insights.

```

# Analysis of Variance
# --------------------
```{r}
library(car)
# ONE WAY ANOVA 
# Question: Do the types of gear affect eh miles per gallon? 
boxplot(mtcars$mpg ~ mtcars$gear, col= rainbow(5))
model1 <- aov(mtcars$mpg ~ mtcars$gear)
summary(model1)

# You can see the ANOVA table and can comment on the ANOVA test result. Comment using F-value and P-value of the test. You need to define Null and Alternative hypothsis as well.

# For help # https://www.r-bloggers.com/2017/08/one-way-anova-in-r/
```

# Linear and Multiple Regression Models:
# --------------------------------------
```{r}
# Linear Regression line formula:
# -------------------------------
attach(mtcars)

fit.LR <- lm(mpg ~ gear, data = mtcars)
summary(fit.LR)  

# This will give details of the model including the 
# correlation, parameters (intercept and slope) along with P-value and Mean sum of squares.  #You need to write comments on these three at least.
#You can add more plots or further analyses/ details/comments as you like. 


# Multiple Regression line formula:
# -------------------------------
#But if you want to fit y dependent on more than 1 independent variables, #then use the following command

fit.MR <- lm(mpg ~ cyl + gear + hp + disp + wt, data = mtcars)
summary(fit.MR)


# beta0 = intercept of the regression line. 
# beta1 = slope of the first independent variable.
# beta2 = slope of the second independent variable. And so on


# Applying test on the coefficicnt

# See Walopole book and check the t-test formula for testing each parameter (intercept, slope and so on). You can see the t-test value and corresponding P-value. You need to comment on the significance of the parameters at 5% level of significance.  


# Is the overall regression model suitable?
# ---------------------------------------
anova(fit.LR)  # Test difference in slopes (joint F-test)
# The F-statistic value will tell whether the result is good or not.

```

